Software for the analysis and visualization of deep mutational scanning data

### 2015-01-08 paper/BMC_Bioinformatics_2015/bmc_bioinformatics_article.tex
Made text changes to LaTex BMC Bioinformatics draft

\section*{Background}
Deep mutational scanning is a high-throughput experimental technique to assess the impacts of mutations on a protein-coding gene~\cite{fowler2014deep}. Figure \ref{fig:DeepMutationalScan} shows a schematic of deep mutational scanning. A gene is mutagenized, and the library of resulting variants is introduced into cells or viruses, which are then subjected to an experimental selection that enriches for functional variants and depletes non-functional ones. Deep sequencing of the variants pre- and post-selection provides information about the functional impacts of mutations. Since the original description of deep mutational scanning by Fowler \textit{et al}~\cite{fowler2010high}, the technique has been applied to a wide range of genes~\cite{traxlmayr2012construction,mclaughlin2012spatial,starita2013activity,melamed2013deep,roscoe2013analyses,firnberg2014comprehensive,bloom2014experimentally,melnikov2014comprehensive,thyagarajan2014inherent,wu2014high,wu2014ns,olson2014comprehensive}, both to measure mutational tolerance given a single selection pressure as in Figure \ref{fig:DeepMutationalScan}A, or to identify mutations that have different effects under alternative selections as in Figure \ref{fig:DeepMutationalScan}B. New techniques to create comprehensive codon-mutant libraries of genes make it possible to profile all single amino-acid mutations~\cite{firnberg2012,jain2013rapid,firnberg2014comprehensive,bloom2014experimentally,melnikov2014comprehensive}, while new techniques for \textit{in vivo} mutagenesis enable deep mutational scanning to be applied across the biological spectrum from viruses and bacteria to human cells~\cite{findlay2014saturation}.

@@

\subsection*{The nature of deep mutational scanning data.}
The data consist of counts of variants pre- and post-selection. The approach presented here treats each site in the gene separately, ignoring epistatic coupling among mutations. This aspect of the approach should not be construed as a suggestion that interactions among mutations are unimportant; indeed, several studies have used deep mutational scanning to examine pairwise epistasis~\cite{olson2014comprehensive,araya2012fundamental,bank2015systematic}, and techniques have been described to obtain linkage between distant sites~\cite{hiatt2010parallel,wu2014hiv}. However, the exploding combinatorics of multiple mutations (a 500-residue protein has only $19 \times 500 \approx 10^4$ single mutants, but $19^2 \times \frac{500 \times 499}{2} \approx 4 \times 10^7$  double mutants and $19^3 \times \frac{500!}{497! \times 3!} \approx 10^{11}$ triple mutants) make it currently plausible to comprehensively characterize only single mutations to all but the shortest genes. Therefore, treating sites independently is not a major limitation for most current datasets; eventually the approach here might be extended to include coupling among mutations.

The data for each site $r$ is characterized by the sequencing \emph{depth} (total number of counts); let $N_r^{\rm{pre}}$, $N_r^{\rm{post}}$, $N_r^{s1}$, and $N_r^{s2}$ denote the depth at $r$ for each of the four libraries in Figure \ref{fig:DeepMutationalScan}. Typical depths for current experiments are $N \sim 10^6$. Denote the counts of character $x$ (characters might be nucleotides, amino acids, or codons) at $r$ as $n_{r,x}^{\rm{pre}}$, $n_{r,x}^{\rm{post}}$, $n_{r,x}^{s1}$, and $n_{r,x}^{s2}$. The values of $n_{r,x}$ for characters $x$ that differ from the wildtype identity $\operatorname{wt}\left(r\right)$ depend on both the depth $N$ and the average per-site mutation rate. Typical experiments introduce about one mutation per gene, so $\overline{\mu} \sim 1/L$ where $L$ is the length of the gene. Therefore, if a 500-codon gene is sequenced at depth $N \sim 10^6$, we expect $N \overline{\mu} \sim 2000$ counts of non-wildtype codons at each site. Since there are 63 mutant codons, the average pre-selection counts for a mutation to a specific $x \ne \operatorname{wt}\left(r\right)$ will be $n_{r,x}^{\rm{pre}} \sim 30$, with counts for most mutations deviating from this average due to biases in creation of the mutant library and randomness in which molecules are sequenced. Counts in the post-selection libraries will further deviate from this average due to selection. Therefore, even at depths $N \sim 10^6$, the actual counts of most mutations will be quite modest.

\subsection*{The challenge of statistical inference from finite counts.}
The goal is to estimate the effects of mutations from changes in their counts after selection. Let $\mu_{r,x}$, $f_{r,x}$, $f_{r,x}^{s1}$, and $f_{r,x}^{s2}$ denote the \emph{true} frequencies of $x$ at $r$ that would be observed for the four libraries in Figure \ref{fig:DeepMutationalScan} if we sampled at infinite depth in both the actual experiment and the sequencing. If we knew these frequencies, we could calculate parameters that reflect the effects of mutations. One parameter that characterizes the effect of mutating $r$ from $\operatorname{wt}\left(r\right)$ to $x$ for the experiment in Figure \ref{fig:DeepMutationalScan}A is the \emph{enrichment ratio}
\begin{equation}
\label{eq:enrichment}
\phi_{r,x} = \frac{f_{r,x} / f_{r,\operatorname{wt}\left(r\right)}}{\mu_{r,x} / \mu_{r,\operatorname{wt}\left(r\right)}}.
\end{equation}
Beneficial mutations have $\phi_{r,x} > 1$, while deleterious ones have $\phi_{r,x} < 1$. A related parameter is the \emph{preference} $\pi_{r,x}$ of $r$ for $x$. At each site, the preferences are simply the enrichment ratios rescaled to sum to one:
\begin{equation}
\label{eq:preference}
\pi_{r,x} = \frac{\phi_{r,x}}{\sum_y \phi_{r,y}} = \frac{f_{r,x} / \mu_{r,x}}{\sum_y f_{r,y} / \mu_{r,y}},
\end{equation}
or equivalently
\begin{equation}
\label{eq:preference2}
f_{r,x} = \frac{\pi_{r,x} \times \mu_{r,x}}{\sum_y \pi_{r,y} \times \mu_{r,y}}.
\end{equation}
The preferences can be intuitively visualized (Figure \ref{fig:DeepMutationalScan}A) and interpreted as the equilibrium frequencies in substitution models for gene evolution~\cite{bloom2014experimentally,bloom2014informed} (after accounting for uneven mutational rates~\cite{yampolsky2005exchangeability,stoltzfus2009climbing}).

Equations \ref{eq:enrichment} and \ref{eq:preference} are in terms of the true frequencies $\mu_{r,x}$, $f_{r,x}$, etc. But in practice, we only observe the counts in the finite sample of sequenced molecules. The computational challenge is to estimate the preferences (or enrichment ratios) from these counts.

@@

This algorithm a slight modification of that in the \textsl{Methods} of \cite{bloom2014experimentally}; here the algorithm is described anew to explain the implementation in \texttt{dms\_tools}.

\subsubsection*{Optional controls to quantify error rates.}
Some sequencing reads that report a mutation may actually reflect an error. Errors can be quantified by sequencing an unmutated gene, so that any counts at $r$ of $x \ne \operatorname{wt}\left(r\right)$ for this control reflect errors. In some cases (e.g. sequencing an RNA virus where the post-selection libraries must be reverse-transcribed), error rates for the pre- and post-selection libraries may differ and so be described by different controls. Let $N_r^{\textrm{errpre}}$ and $N_r^{\rm{errpost}}$ be the depth and $n_{r,x}^{\rm{errpre}}$ and $n_{r,x}^{\rm{errpost}}$ be the counts of $x$ in the pre-selection and post-selection error controls, respectively. Define $\epsilon_{r,x}$ and $\rho_{r,x}$ to be the true frequencies of errors at $r$ from $\operatorname{wt}\left(r\right)$ to $x$ in the pre- and post-selection controls, respectively.

@@

\subsubsection*{Priors over the unknown parameters.}
We specify Dirichlet priors over the parameter vectors:
\begin{gather}
\Pr\left(\boldsymbol{\mathbf{\pi_r}}\right) = \operatorname{Dirichlet}\left(\boldsymbol{\mathbf{\pi_r}}; \alpha_{\pi} \times \mathbf{1}\right)   \label{eq:pr_pir} \\
\Pr\left(\boldsymbol{\mathbf{\mu_r}}\right) = \operatorname{Dirichlet}\left(\boldsymbol{\mathbf{\mu_r}}; \alpha_{\mu} \times \mathcal{N}_x \times \boldsymbol{\mathbf{a_{r,\mu}}}\right) \label{eq:pr_mur} \\
\Pr\left(\boldsymbol{\mathbf{\epsilon_r}}\right) = \operatorname{Dirichlet}\left(\boldsymbol{\mathbf{\epsilon_r}}; \alpha_{\epsilon} \times \mathcal{N}_x \times \boldsymbol{\mathbf{a_{r,\epsilon}}}\right) \label{eq:pr_epsilonr} \\
\Pr\left(\boldsymbol{\mathbf{\rho_r}}\right) = \operatorname{Dirichlet}\left(\boldsymbol{\mathbf{\rho_r}}; \alpha_{\rho} \times \mathcal{N}_x \times
 \boldsymbol{\mathbf{a_{r,\rho}}}\right) \label{eq:pr_rhor}
\end{gather}
where $\mathbf{1}$ is a vector of ones, $\mathcal{N}_x$ is the number of characters (64 for codons, 20 for amino acids, 4 for nucleotides), the $\alpha$'s are scalar concentration parameters $> 0$ (by default \texttt{dms\_tools} sets the $\alpha$'s to one). For codons, the error rate depends on the number of nucleotides being changed. The average error rates  $\overline{\epsilon_m}$ and $\overline{\rho_m}$ for $m$ nucleotide changes are estimated as
\begin{gather}
\overline{\epsilon_m} = \frac{1}{L}\sum\limits_r \frac{1}{N_r^{\textrm{errpre}}}\sum\limits_{x} n_{r,x}^{\textrm{errpre}}\times \delta_{m,D_{x,\operatorname{wt}\left(r\right)}} \label{eq:avgepsilonm} \\
\overline{\rho_m} = \frac{1}{L}\sum\limits_r \frac{1}{N_r^{\textrm{errpost}}}\sum\limits_{x} n_{r,x}^{\textrm{errpost}}\times \delta_{m,D_{x,\operatorname{wt}\left(r\right)}} \label{eq:avgrhom}
\end{gather}
where $D_{x,\operatorname{wt}\left(r\right)}$ is the number of nucleotide differences between $x$ and $\operatorname{wt}\left(r\right)$. Given these definitions, 
\begin{gather}
\boldsymbol{\mathbf{a_{r,\epsilon}}} = \left(\cdots, \sum\limits_m \frac{\overline{\epsilon}_{m}}{\mathcal{C}_m} \times \delta_{m,D_{x,\operatorname{wt}\left(r\right)}},\cdots\right) \label{eq:arepsilon} \\
\boldsymbol{\mathbf{a_{r,\rho}}} = \left(\cdots, \sum\limits_m \frac{\overline{\rho}_{m}}{\mathcal{C}_m} \times \delta_{m,D_{x,\operatorname{wt}\left(r\right)}}, \cdots\right) \label{eq:arrho} 
\end{gather}
where $\mathcal{C}_m$ is the number of mutant characters with $m$ changes relative to wildtype (for nucleotides $\mathcal{C}_0 = 1$ and $\mathcal{C}_1 = 3$; for codons $\mathcal{C}_0 = 1$, $\mathcal{C}_1 = 9$, $\mathcal{C}_2 = \mathcal{C}_3 = 27$).

Our prior assumption is that the mutagenesis introduces all mutant characters at equal frequency (this assumption is only plausible for codons if the mutagenesis is at the codon level as in~\cite{firnberg2012,jain2013rapid,firnberg2014comprehensive,bloom2014experimentally,melnikov2014comprehensive}). The per-site mutagenesis rate is estimated as
\begin{equation}
\label{eq:avgmu}
\overline{\mu} = \left(\frac{1}{L}\sum\limits_r \frac{1}{N_r^{\textrm{pre}}}\sum\limits_{x\ne \operatorname{wt}\left(r\right)} n_{r,x}^{\textrm{pre}}\right) - \sum\limits_{m \ge 1} \overline{\epsilon_m},
\end{equation}
so that
\begin{equation}
\boldsymbol{\mathbf{a_{r,\mu}}} = \left(\cdots, \frac{\overline{\mu}}{\mathcal{N}_x - 1} + \delta_{x,\operatorname{wt}\left(r\right)} \times \left[1 - \overline{\mu}\right] ,\cdots\right). \label{eq:armu}
\end{equation}

@@

\subsection*{Inferring preferences with \texttt{dms\_tools}.}
\subsubsection*{Application to actual datasets.}
Figures \ref{fig:MelnikovPrefs} and \ref{fig:ThyagarajanPrefs} illustrate application of \texttt{dms\_tools} to two existing datasets~\cite{melnikov2014comprehensive,thyagarajan2014inherent}. As these figures show, the \texttt{dms\_inferprefs} and \texttt{dms\_logoplot} programs infer and visualize the preferences with a few simple commands. Error controls can be included when available (they are not for Figure \ref{fig:MelnikovPrefs}, but are for Figure \ref{fig:ThyagarajanPrefs}). The runtime for the MCMC depends on the gene length and character type (codons are slowest, nucleotides fastest) -- but if the inference is parallelized across multiple CPUs (using the \texttt{--ncpus} option of \texttt{dms\_inferprefs}), the inference should take no more than a few hours. As shown in Figures \ref{fig:MelnikovPrefs} and \ref{fig:ThyagarajanPrefs}, the visualizations can overlay information about protein structure onto the preferences.

@@

\subsection*{Algorithm to infer differential preferences.}
As shown in Figure \ref{fig:DeepMutationalScan}B, a useful extension to the experiment in Figure \ref{fig:DeepMutationalScan}A is to subject the functional variants to two different selection pressures to identify mutations favored by one pressure versus the other. While this experiment could in principle by analyzed by simply comparing the initial unselected mutants to the final variants after the two alternative selections, this approach is non-ideal. In experiments like Figure \ref{fig:DeepMutationalScan}A, many mutations are enriched or depleted to some extent by selection, since a large fraction of random mutations affect protein function~\cite{shortle1985,rennell1991,shafikhani1997,guo2004,bloom2005}. Therefore, the assumption that all mutations are equally tolerated (i.e. the preferences for a site are all equal, or the enrichment ratios are all one) is not a plausible null hypothesis for Figure \ref{fig:DeepMutationalScan}A, and so \texttt{dms\_tools} simply infers the preferences given a uniform Dirichlet prior rather than trying to pinpoint some subset of sites with unequal preferences.

But in Figure \ref{fig:DeepMutationalScan}B, the assumption that most mutations will be similarly selected is a plausible null hypothesis, since only a small subset of mutations will have markedly different effects under two related selections. Therefore, \texttt{dms\_tools} uses a different algorithm to infer the differential preferences under the two selections. This algorithm combines a prior that mildly favors differential preferences of zero with a likelihood-based analysis of the mutation counts to estimate posterior probabilities over the differential preferences.

\subsubsection*{Definition of the differential preferences.}
Given an experiment like Figure \ref{fig:DeepMutationalScan}B, let $f_{r,x}^{\textrm{start}}$ be the true frequency of character $x$ at site $r$ in the starting library (equivalent to the frequency $f_{r,x}^{\textrm{post}}$ in the figure), and let $f_{r,x}^{s1}$ and $f_{r,x}^{s2}$ be the frequencies after selections $s1$ and $s2$, respectively. The differential preference $\Delta\pi_{r,x}$ for $x$ at $r$ in $s2$ versus $s1$ is defined by:
\begin{gather}
\label{eq:controlpi}
f_{r,x}^{s1} = \frac{f_{r,x}^{\textrm{start}} \times \pi_{r,x}^{s1}}{\sum_y f_{r,y}^{\textrm{start}} \times \pi_{r,y}^{s1}} \\
\label{eq:deltapi}
f_{r,x}^{s2} = \frac{f_{r,x}^{\textrm{start}} \times \left(\pi_{r,x}^{s1} + \Delta\pi_{r,x}\right)}{\sum_y f_{r,y}^{\textrm{start}} \times \left(\pi_{r,y}^{s1} + \Delta\pi_{r,y}\right)}
\end{gather}
where $\pi_{r,x}^{s1}$ is the ``control preference'' and is treated as a nuisance parameter, and we have the constraints
\begin{gather}
\label{eq:diffprefsum}
0 = \sum\limits_x \Delta\pi_{r,x} \\
\label{eq:diffprefpluspref}
0 \le \pi_{r,x}^{s1} + \Delta\pi_{r,x} \le 1.
\end{gather}
If there is no difference in the effect of $x$ at $r$ between selections $s1$ and $s2$, then $\Delta\pi_{r,x} = 0$. If $x$ at $r$ is more preferred by $s2$ than $s1$, then $\Delta\pi_{r,x} > 0$; conversely if $x$ at $r$ is more preferred by $s1$ than $s2$, then $\Delta\pi_{r,x} < 0$.

>>>>

\section*{Background}
Deep mutational scanning is a high-throughput experimental technique to assess the impacts of mutations on a protein-coding gene~\cite{fowler2014deep}. Figure \ref{fig:DeepMutationalScan} shows a schematic of deep mutational scanning. A gene is mutagenized, and the library of resulting variants is introduced into cells or viruses, which are then subjected to an experimental selection that enriches for functional variants and depletes non-functional ones. Deep sequencing of the variants pre- and post-selection provides information about the functional impacts of mutations. Since the original description of deep mutational scanning by Fowler \textit{et al}~\cite{fowler2010high}, the technique has been applied to a wide range of genes~\cite{traxlmayr2012construction,mclaughlin2012spatial,starita2013activity,melamed2013deep,roscoe2013analyses,firnberg2014comprehensive,bloom2014experimentally,melnikov2014comprehensive,thyagarajan2014inherent,wu2014high,wu2014ns,olson2014comprehensive}, both to measure mutational tolerance given a single selection pressure as in Figure \ref{fig:DeepMutationalScan}A, or to identify mutations that have different effects under alternative selections as in Figure \ref{fig:DeepMutationalScan}B. New techniques to create comprehensive codon-mutant libraries of genes make it possible to profile all single amino-acid mutations~\cite{firnberg2012,jain2013rapid,firnberg2014comprehensive,bloom2014experimentally,melnikov2014comprehensive}, while new techniques for targeted mutagenesis of mammalian genomes enable deep mutational scanning to be applied across the biological spectrum from viruses and bacteria to human cells~\cite{findlay2014saturation}.

@@

\subsection*{The nature of deep mutational scanning data.}
The data consist of counts of variants pre- and post-selection. The approach presented here treats each site in the gene separately, ignoring epistatic coupling among mutations. This aspect of the approach should not be construed as a suggestion that interactions among mutations are unimportant; indeed, several studies have used deep mutational scanning to examine pairwise epistasis~\cite{olson2014comprehensive,araya2012fundamental,bank2015systematic}, and techniques have been described to obtain linkage between distant sites~\cite{hiatt2010parallel,wu2014hiv}. However, the exploding combinatorics of multiple mutations (a 500-residue protein has only $19 \times 500 \approx 10^4$ single mutants, but $19^2 \times \frac{500 \times 499}{2} \approx 4 \times 10^7$  double mutants and $19^3 \times \frac{500!}{497! \times 3!} \approx 10^{11}$ triple mutants) make it currently plausible to comprehensively characterize only single mutations to all but the shortest genes. Treating sites independently is therefore not a major limitation for most current datasets. Eventually the approach here might be extended to include coupling among mutations.

The data for each site $r$ is characterized by the sequencing \emph{depth} (total number of counts); let $N_r^{\rm{pre}}$, $N_r^{\rm{post}}$, $N_r^{s1}$, and $N_r^{s2}$ denote the depth at $r$ for each of the four libraries in Figure \ref{fig:DeepMutationalScan} (pre-selection, post-selection, selection $s1$, and selection $s2$). Typical depths for current experiments are $N \sim 10^6$. Denote the counts of character $x$ (characters might be nucleotides, amino acids, or codons) at $r$ as $n_{r,x}^{\rm{pre}}$, $n_{r,x}^{\rm{post}}$, $n_{r,x}^{s1}$, and $n_{r,x}^{s2}$. The values of $n_{r,x}$ for characters $x$ that differ from the wildtype identity $\operatorname{wt}\left(r\right)$ depend on both the depth $N$ and the average per-site mutation rate. Typical experiments introduce about one mutation per gene, so the average mutation rate is $\overline{\mu} \sim 1/L$ where $L$ is the length of the gene. Therefore, if a 500-codon gene is sequenced at depth $N \sim 10^6$, we expect $N \overline{\mu} \sim 2000$ counts of non-wildtype codons at each site. Since there are 63 mutant codons, the average pre-selection counts for a mutation to a specific $x \ne \operatorname{wt}\left(r\right)$ will be $n_{r,x}^{\rm{pre}} \sim 30$, with counts for most mutations deviating from this average due to biases in creation of the mutant library and randomness in which molecules are sequenced. Counts in the post-selection libraries will further deviate from this average due to selection. Therefore, even at depths $N \sim 10^6$, the actual counts of most mutations will be quite modest.

\subsection*{The goal: inferring site-specific amino-acid preferences.}
The goal is to estimate the effects of mutations from changes in their counts after selection. Let $\mu_{r,x}$, $f_{r,x}$, $f_{r,x}^{s1}$, and $f_{r,x}^{s2}$ denote the \emph{true} frequencies of $x$ at $r$ that would be observed for the four libraries in Figure \ref{fig:DeepMutationalScan} if we sampled at infinite depth in both the actual experiment and the sequencing. If we knew these frequencies, we could calculate parameters that reflect the effects of mutations. One parameter that characterizes the effect of mutating $r$ from $\operatorname{wt}\left(r\right)$ to $x$ for the experiment in Figure \ref{fig:DeepMutationalScan}A is the \emph{enrichment ratio}, which is the relative frequency of mutations to $x$ after selection versus before selection:
\begin{equation}
\label{eq:enrichment}
\phi_{r,x} = \frac{f_{r,x} / f_{r,\operatorname{wt}\left(r\right)}}{\mu_{r,x} / \mu_{r,\operatorname{wt}\left(r\right)}}.
\end{equation}
Beneficial mutations have $\phi_{r,x} > 1$, while deleterious ones have $\phi_{r,x} < 1$. A related parameter is the \emph{preference} $\pi_{r,x}$ of $r$ for $x$. At each site, the preferences are simply the enrichment ratios rescaled to sum to one:
\begin{equation}
\label{eq:preference}
\pi_{r,x} = \frac{\phi_{r,x}}{\sum_y \phi_{r,y}} = \frac{f_{r,x} / \mu_{r,x}}{\sum_y f_{r,y} / \mu_{r,y}},
\end{equation}
or equivalently
\begin{equation}
\label{eq:preference2}
f_{r,x} = \frac{\pi_{r,x} \times \mu_{r,x}}{\sum_y \pi_{r,y} \times \mu_{r,y}},
\end{equation}
where $y$ is summed over all character identities (all nucleotides, codons, or amino acids).
The preferences can be intuitively visualized (Figure \ref{fig:DeepMutationalScan}A) and interpreted as the equilibrium frequencies in substitution models for gene evolution~\cite{bloom2014experimentally,bloom2014informed} (after accounting for uneven mutational rates~\cite{yampolsky2005exchangeability,stoltzfus2009climbing}).

\subsection*{The challenge of statistical inference from finite counts.}
Equations \ref{eq:enrichment} and \ref{eq:preference} are in terms of the true frequencies $\mu_{r,x}$, $f_{r,x}$, etc. But in practice, we only observe the counts in the finite sample of sequenced molecules. The computational challenge is to estimate the preferences (or enrichment ratios) from these counts.

@@

This algorithm is a slight modification of that in the \textsl{Methods} of \cite{bloom2014experimentally}; here the algorithm is described anew to explain the implementation in \texttt{dms\_tools}.

\subsubsection*{Optional controls to quantify error rates.}
Some sequencing reads that report a mutation may actually reflect an error introduced during sequencing or PCR rather than an actual mutation that experienced selection. Errors can be quantified by sequencing an unmutated gene, so that any counts at $r$ of $x \ne \operatorname{wt}\left(r\right)$ for this control reflect errors. In some cases (e.g. sequencing an RNA virus where the post-selection libraries must be reverse-transcribed), error rates for the pre- and post-selection libraries may differ and so be described by different controls. Let $N_r^{\textrm{errpre}}$ and $N_r^{\rm{errpost}}$ be the depth and $n_{r,x}^{\rm{errpre}}$ and $n_{r,x}^{\rm{errpost}}$ be the counts of $x$ in the pre-selection and post-selection error controls, respectively. Define $\epsilon_{r,x}$ and $\rho_{r,x}$ to be the true frequencies of errors at $r$ from $\operatorname{wt}\left(r\right)$ to $x$ in the pre- and post-selection controls, respectively.

@@

\subsubsection*{Priors over the unknown parameters.}
We specify Dirichlet priors over the parameter vectors:
\begin{gather}
\Pr\left(\boldsymbol{\mathbf{\pi_r}}\right) = \operatorname{Dirichlet}\left(\boldsymbol{\mathbf{\pi_r}}; \alpha_{\pi} \times \mathbf{1}\right)   \label{eq:pr_pir} \\
\Pr\left(\boldsymbol{\mathbf{\mu_r}}\right) = \operatorname{Dirichlet}\left(\boldsymbol{\mathbf{\mu_r}}; \alpha_{\mu} \times \mathcal{N}_x \times \boldsymbol{\mathbf{a_{r,\mu}}}\right) \label{eq:pr_mur} \\
\Pr\left(\boldsymbol{\mathbf{\epsilon_r}}\right) = \operatorname{Dirichlet}\left(\boldsymbol{\mathbf{\epsilon_r}}; \alpha_{\epsilon} \times \mathcal{N}_x \times \boldsymbol{\mathbf{a_{r,\epsilon}}}\right) \label{eq:pr_epsilonr} \\
\Pr\left(\boldsymbol{\mathbf{\rho_r}}\right) = \operatorname{Dirichlet}\left(\boldsymbol{\mathbf{\rho_r}}; \alpha_{\rho} \times \mathcal{N}_x \times
 \boldsymbol{\mathbf{a_{r,\rho}}}\right) \label{eq:pr_rhor}
\end{gather}
where $\mathbf{1}$ is a vector of ones, $\mathcal{N}_x$ is the number of characters (64 for codons, 20 for amino acids, 4 for nucleotides), the $\alpha$'s are scalar concentration parameters $> 0$ (by default \texttt{dms\_tools} sets the $\alpha$'s to one). For codons, the error rate depends on the number of nucleotides being changed. The average error rates $\overline{\epsilon_m}$ and $\overline{\rho_m}$ for codon mutations with $m$ nucleotide changes are estimated as
\begin{gather}
\overline{\epsilon_m} = \frac{1}{L}\sum\limits_r \frac{1}{N_r^{\textrm{errpre}}}\sum\limits_{x} n_{r,x}^{\textrm{errpre}}\times \delta_{m,D_{x,\operatorname{wt}\left(r\right)}} \label{eq:avgepsilonm} \\
\overline{\rho_m} = \frac{1}{L}\sum\limits_r \frac{1}{N_r^{\textrm{errpost}}}\sum\limits_{x} n_{r,x}^{\textrm{errpost}}\times \delta_{m,D_{x,\operatorname{wt}\left(r\right)}} \label{eq:avgrhom}
\end{gather}
where $D_{x,\operatorname{wt}\left(r\right)}$ is the number of nucleotide differences between $x$ and $\operatorname{wt}\left(r\right)$. Given these definitions, 
\begin{gather}
\boldsymbol{\mathbf{a_{r,\epsilon}}} = \left(\cdots, \sum\limits_m \frac{\overline{\epsilon}_{m}}{\mathcal{C}_m} \times \delta_{m,D_{x,\operatorname{wt}\left(r\right)}},\cdots\right) \label{eq:arepsilon} \\
\boldsymbol{\mathbf{a_{r,\rho}}} = \left(\cdots, \sum\limits_m \frac{\overline{\rho}_{m}}{\mathcal{C}_m} \times \delta_{m,D_{x,\operatorname{wt}\left(r\right)}}, \cdots\right) \label{eq:arrho} 
\end{gather}
where $\mathcal{C}_m$ is the number of mutant characters with $m$ changes relative to wildtype (for nucleotides $\mathcal{C}_0 = 1$ and $\mathcal{C}_1 = 3$; for codons $\mathcal{C}_0 = 1$, $\mathcal{C}_1 = 9$, $\mathcal{C}_2 = \mathcal{C}_3 = 27$).

Our prior assumption is that the mutagenesis introduces all mutant characters at equal frequency (this assumption is only plausible for codons if the mutagenesis is at the codon level as in~\cite{firnberg2012,jain2013rapid,firnberg2014comprehensive,bloom2014experimentally,melnikov2014comprehensive}; if mutations are made at the nucleotide level such as by error-prone PCR then characters should be defined as nucleotides). The average per-site mutagenesis rate is estimated as
\begin{equation}
\label{eq:avgmu}
\overline{\mu} = \left(\frac{1}{L}\sum\limits_r \frac{1}{N_r^{\textrm{pre}}}\sum\limits_{x\ne \operatorname{wt}\left(r\right)} n_{r,x}^{\textrm{pre}}\right) - \sum\limits_{m \ge 1} \overline{\epsilon_m},
\end{equation}
so that
\begin{equation}
\boldsymbol{\mathbf{a_{r,\mu}}} = \left(\cdots, \frac{\overline{\mu}}{\mathcal{N}_x - 1} + \delta_{x,\operatorname{wt}\left(r\right)} \times \left[1 - \overline{\mu}\right] ,\cdots\right). \label{eq:armu}
\end{equation}

@@

\subsection*{Inferring preferences with \texttt{dms\_tools}.}

\subsubsection*{Application to actual datasets.}
Figures \ref{fig:MelnikovPrefs} and \ref{fig:ThyagarajanPrefs} illustrate application of \texttt{dms\_tools} to two existing datasets~\cite{melnikov2014comprehensive,thyagarajan2014inherent}. The programs require as input only simple text files listing the counts of each character identity at each site. As the figures show, the \texttt{dms\_inferprefs} and \texttt{dms\_logoplot} programs can process these input files to infer and visualize the preferences with a few simple commands. Error controls can be included when available (they are not for Figure \ref{fig:MelnikovPrefs}, but are for Figure \ref{fig:ThyagarajanPrefs}). The runtime for the MCMC depends on the gene length and character type (codons are slowest, nucleotides fastest) -- but if the inference is parallelized across multiple CPUs (using the \texttt{--ncpus} option of \texttt{dms\_inferprefs}), the inference should take no more than a few hours. As shown in Figures \ref{fig:MelnikovPrefs} and \ref{fig:ThyagarajanPrefs}, the visualizations can overlay information about protein structure onto the preferences.

@@

\subsection*{Algorithm to infer differential preferences.}
As shown in Figure \ref{fig:DeepMutationalScan}B, a useful extension to the experiment in Figure \ref{fig:DeepMutationalScan}A is to subject the functional variants to two different selection pressures to identify mutations favored by one pressure versus the other. While this experiment could in principle by analyzed by simply comparing the initial unselected mutants to the final variants after the two alternative selections, this approach is non-ideal. In experiments like Figure \ref{fig:DeepMutationalScan}A, many mutations are enriched or depleted to some extent by selection, since a large fraction of random mutations affect protein function~\cite{shortle1985,rennell1991,shafikhani1997,guo2004,bloom2005}. Therefore, the assumption that all mutations are equally tolerated (i.e. the preferences for a site are all equal, or the enrichment ratios are all one) is not a plausible null hypothesis for Figure \ref{fig:DeepMutationalScan}A. For this reason, \texttt{dms\_tools} simply infers the preferences given a uniform Dirichlet prior rather than trying to pinpoint some subset of sites with unequal preferences.

But in Figure \ref{fig:DeepMutationalScan}B, the assumption that most mutations will be similarly selected is a plausible null hypothesis, since we expect to alternative selections to have markedly different effects on only a small subset of mutations (typically, major constraints related to protein folding and stability will be relatively conserved across different selections on the same protein). Therefore, \texttt{dms\_tools} uses a different algorithm to infer the differential preferences under the two selections. This algorithm combines a prior that mildly favors differential preferences of zero with a likelihood-based analysis of the mutation counts to estimate posterior probabilities over the differential preferences.

\subsubsection*{Definition of the differential preferences.}
Given an experiment like Figure \ref{fig:DeepMutationalScan}B, let $f_{r,x}^{\textrm{start}}$ be the true frequency of character $x$ at site $r$ in the starting library (equivalent to the frequency $f_{r,x}^{\textrm{post}}$ in the figure), and let $f_{r,x}^{s1}$ and $f_{r,x}^{s2}$ be the frequencies after selections $s1$ and $s2$, respectively. The differential preference $\Delta\pi_{r,x}$ for $x$ at $r$ in $s2$ versus $s1$ is defined by:
\begin{gather}
\label{eq:controlpi}
f_{r,x}^{s1} = \frac{f_{r,x}^{\textrm{start}} \times \pi_{r,x}^{s1}}{\sum_y f_{r,y}^{\textrm{start}} \times \pi_{r,y}^{s1}} \\
\label{eq:deltapi}
f_{r,x}^{s2} = \frac{f_{r,x}^{\textrm{start}} \times \left(\pi_{r,x}^{s1} + \Delta\pi_{r,x}\right)}{\sum_y f_{r,y}^{\textrm{start}} \times \left(\pi_{r,y}^{s1} + \Delta\pi_{r,y}\right)}
\end{gather}
where $\pi_{r,x}^{s1}$ is the ``control preference'' and is treated as a nuisance parameter, and we have the constraints
\begin{gather}
\label{eq:diffprefsum}
0 = \sum\limits_x \Delta\pi_{r,x} \\
\label{eq:diffprefpluspref}
0 \le \pi_{r,x}^{s1} + \Delta\pi_{r,x} \le 1.
\end{gather}
If there is no difference in the effect of $x$ at $r$ between selections $s1$ and $s2$, then $\Delta\pi_{r,x} = 0$. If $x$ at $r$ is more preferred by $s2$ than $s1$, then $\Delta\pi_{r,x} > 0$; conversely if $x$ at $r$ is more preferred by $s1$ than $s2$, then $\Delta\pi_{r,x} < 0$ (see Figure \ref{fig:MelnikovSimulatedDiffPrefs}A).
